---
title: "R Functions"
author: "Kylie Ariel Bemis"
date: "5/27/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions in R

*To understand computations in R, two slogans are helpful:*

- *Everything that exists is an object*
- *Everything that happens is a function call*

-– John Chambers, creator of S

## "Everything that exists is an object"

Consider these simple lines of code:

```{r}
x <- 2
y <- 3
x + y
```

What actually happens when you run `x + y`?

## "Everything that exists is an object"

```{r}
sexp <- quote(x + y)
sexp
```

What type of object is `sexp`?

```{r}
typeof(sexp)
```

`typeof` returns an object's **base type**.

## "Everything that exists is an object"

R code itself is a language object that can be manipulated, changed, and evaluated.

```{r}
sexp
eval(sexp)
sexp[[1]] <- quote(`*`)
sexp
eval(sexp)
```

## "Everything that exists is an object"

```{r}
sexp
as.list(sexp)
```

Why does the `*` come first in the object? (Hint: see next slide.)

## "Everything that happens is a function call"

Reconsider:

```{r}
x <- 2
y <- 3
x + y
```

This is the same as doing:

```{r}
`<-`(x, 2)
`<-`(y, 3)
`+`(x, y)
```

## "Everything that happens is a function call"

Why does this work?

```{r}
`<-`(x, 2)
`<-`(y, 3)
`+`(x, y)
```

In R, addition is just a function for which we commonly use the **infix** notation, but using its **prefix** notation works just as well, and is how functions are internally stored in R.

(This is why we accessed `*` using `sexp[[1]]` instead of `sexp[[2]]`.)

## Functions in R

Functions are first-class citizens in R. They are objects that can be passed around and manipulated like any other object.

Functions in R have three key characteristics:

- `body` -- the code inside the function
- `formals` -- a list of arguments used to call the function
- `environment` -- where to find the function's variables

You provide `body` and `formals` when defining a function. The `environment` is defined automatically by **where you are** when you define it.

## Functions in R (cont'd)

```{r}
add <- function(x, y) x + y
body(add)
formals(add)
environment(add)
```

## Exception: primitive functions

Some low-level "primitive" functions defined by the core R team are exceptions to this, and call C code immediately upon being called. `+` is actually a primitive function.

```{r}
`+`
```

Primitive functions only exist in the `base` R package and can only be created by the core R team, so we won't discuss them any further beyond acknowledging their existence.

## Defining a function in R

Functions in R are defined by the `function` function:

- The arguments you provide to `function` become the formal arguments of your function
- An expression follows that becomes the body of the function
- Your current environment becomes the environment for the function

```{r}
add <- function(x, y) x + y
add
```

Note that while you can explicitly return values with the `return` function, most R functions simply return the value of the last evaluated expression in the body. In our `add` function above, that is simply `x + y`.

## Anonymous functions

Note that we don't actually have to assign the function to a variable to use it.

```{r}
(function(x, y) x + y)
(function(x, y) x + y)(1, 2)
```

This is called an **anonymous function**. Anonymous functions are useful when using functions like `lapply`, `sapply`, and `purrr::map`.

## Lexical scoping

How does a function find values for the variables in its `body`?

```{r}
add_1 <- function(x) x + 1
add_1
add_1(1)
add_y <- function(x) x + y
add_y
```

It is clear what `add_1` does. But what will `add_y` do to find `y`?

## Lexical scoping

\small
Functions capture the environment in which they were created, and have access to all variables in the environment.

Because we created `add_y` in the global environment, that means it has access to all variables in the global environment.

We simply need to define a `y` variable in the global environment.

```{r}
add_y
environment(add_y)
y <- 2
add_y(1)
```

Why would we want to do something like this?
\normalsize

## Functionals

Suppose we wish to create a function that allows a user to add some number `val` to any number, but we don't know what `val` will be. We can simply create that function once we know what `val` is!

A function that returns a function like this (or takes a function as an argument) is called a **functional**. Functionals are common in R, most notably in functions like `lapply`, `sapply`, and `purrr::map`.

```{r}
add_val <- function(val) {
  function(x) x + val
}
add_10 <- add_val(10)
add_10(1)
```

What happened here?

## Lexical scoping (cont'd)

When a function is called in R, the following happens:

- A new, temporary environment is created
- Any formal arguments of the function are assigned to the temporary environment
- The temporary environment's *parent environment* (or "enclosing" environment) is the *function’s environment*
- The function is evaluated in this temporary environment
- When a variable name is encountered, R searches the current (temporary) environment, then its parent environment (the function’s environment), then its parent’s parent environment, and so on, until the variable is found

## Lexical scoping and closures

```{r}
add_val <- function(val) {
  function(x) x + val
}
add_10 <- add_val(10)
add_10
```

When we evaluate `add_val`, it creates a temporary environment and assigns `val` into it. It then returns a new function whose environment *is* the "temporary" environment created by evaluating `add_val`, which is where `val` can be found. Now our new function `add_10` always has access to `val` (which in our example is 10).

When a function is stored together with its environment like this, it's called a **closure**.

## Functions (cont'd)

There is a lot more that can be done with functions in R, but we don't need most of those things. See the __[Advanced R](http://adv-r.had.co.nz)__ textbook chapter on functions for more information on:

- Default arguments and missing arguments

- Lazy evaluation

- Special calls

    - Infix functions
    
    - Replacement functions

and more.
